\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-15 analysis/paper/paper.tex
update CS3

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}

@@

\hline

@@

\hline
\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule
\hline

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
R+                                event = "DeadOfPrCa", exposure = "ScrArm")
R> inherits(pt_object, "popTime")
\end{CodeInput}

@@

\begin{CodeInput}
R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
R+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
R+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
R+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
R> 
R> fits <- lapply(fmla, function(i) {
R+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
R+ })
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
R> new_time <- seq(0,14,0.1)
R> 
R> risks <- lapply(fits, function(i) {
R>   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
R> })
\end{CodeInput}
\end{CodeChunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted above, the usual asymptotic results hold for
likelihood ratio tests built using case-base sampling models. Therefore,
we can easily test the null hypothesis that the Weibull model is just as
good as the larger (in terms of number of parameters) spline model:

@@

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
#> 1     54537       5801                         
#> 2     54535       5785  2     15.7    4e-04 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase::fitSmoothHazard & survival::survreg\\
\midrule
Exponential & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.79 (0.66, 0.94) & --\\
\bottomrule
\multicolumn{3}{l}{Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
\end{tabular}
\end{table}

@@

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
prostate cancer hazard ratio and 95\textbackslash\% confidence interval
for screening vs.~control group as a function of time using the
\code{plot} method for objects of class \code{singleEventCB}:

@@

\begin{CodeInput}
R> plot(fit_inter, type = "hr", newdata = new_data,
R>      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+   data = bmtcrr
R+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
R+                         const(Phase) + const(Source) + const(Age),
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. For more information
about this dataset, the reader is encouraged to look at the
documentation in our package. A description of the variables can be
found in Table \ref{tab:support1}, and a breakdown of each categorical
variable appears in Table \ref{tab:support2}.
\begin{tabular}{ccccc}
\hline
\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
\hline
age           & Age                                      &                 & double           & 0            \\
death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
sex           &                                          & 2               & integer          & 0            \\
slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
d.time        & Days of Follow-Up                        &                 & double           & 0            \\
dzgroup       &                                          & 8               & integer          & 0            \\
dzclass       &                                          & 4               & integer          & 0            \\
num.co        & number of comorbidities                  &                 & double           & 0            \\
edu           & Years of Education                       &                 & double           & 202          \\
income        &                                          & 4               & integer          & 349          \\
scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
race          &                                          & 5               & integer          & 5            \\
sps             & support physiology score day 3           &                   & double           & 1            \\
aps           & APS III no coma, imp bun,uout for ph1,D3 &                 & double           & 1            \\
hday            & Day in Hospital at Study Admit             &                 & double           & 0            \\ 
diabetes      & Diabetes (Com 27-28, Dx 73)              &                 & double           & 0            \\
dementia      & Dementia (Comorbidity 6)                 &                 & double           & 0            \\
ca              & Cancer State                             & 3               & integer          & 0            \\
meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
bili          & Bilirubin Day 3                          &                 & double           & 297          \\
crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
sod           & Serum sodium Day 3                       &                 & double           & 0            \\
ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
glucose       & Glucose Day 3                            &                 & double           & 470          \\
bun           & BUN Day 3                                &                 & double           & 455          \\
urine         & Urine Output Day 3                       &                 & double           & 517          \\
adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
\end{tabular}
\caption{Description of each variable in the SUPPORT dataset.}
\label{tab:support1}

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
\label{tab:support2}
\end{table}

In the original study, the authors developed two scores for predicting
the outcome. These scores are available as part of the data: SUPPORT day
3 physiology score (\code{sps}) and APACHE III day 3 physiology score
(\code{aps}). Using these scores, we will explore three Cox regression
scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
We will compare the three models using a time-dependent version of the
classical Brier score that is adjusted for censoring
\citep{graf1999ass}. The Brier score can be used to assess both
discrimination and calibration in a single model. Moreover, it provides
a metric that can be used to compare parametric and semi-parametric
survival models. To account for overfitting, we will use 95\% of the
observations to train our models, while the remaining 5\% will be used
to compute absolute risk curves and the Brier scores.

In Figure\ref{fig:brier1} (A), we can see that all Cox models have
similar average survival curves. On the other hand, Figure
\ref{fig:brier1} (B) shows that the full model has the lowest Brier
score overall. For this reason, we will use this model in our comparison
between regularized Cox regression, Kaplan-Meier estimation, and the
approach from \pkg{casebase}.

{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/brierpt1-1}

\caption{\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve. (A) Cumulative Incidence curves. For a given model, the plotted curve is the average curve of each individual in the study. Each follow-up time seen in the study was used. There are minimal differences between the curves. APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better. The full model performs best overall.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

For our penalized model, we opt for the natural log of time; recall that
this corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the function \code{fitSmoothHazard}. In order to use this interface,
we first create a matrix \code{y} containing the time and event
variables, and a matrix \code{x} containing all remaining covariates we
would like to include in the model space. We used lasso penalization by
setting \code{alpha = 1}. Moreover, we use the argument
\code{penalty.factor} to force the time variable into the final model.

@@

\begin{CodeInput}
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
R+   family = "glmnet",
R+   time = "d.time", event = "death",
R+   formula_time = ~ log(d.time), alpha = 1,
R+   ratio = 10, standardize = TRUE,
R+   penalty.factor = c(0, rep(1, ncol(x)))
R+ )
\end{CodeInput}
\end{CodeChunk}

Next, we can use the object \code{cbFull} to estimate the absolute risk
curve on the test data. We first create a matrix \code{newx} containing
the relevant variables. Next, using \code{times}, we specify the time
points at which the cumulative incidence will be computed. Finally,
using the parameter \code{s = "lambda.1se"}, we specify at which value
of the tuning parameter we want to retrieve the coefficient estimates.

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newdata parameter
R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = test)[, -c(1)]
R> times <- sort(unique(test$d.time))
R> 
R> abCbFull <- casebase::absoluteRisk(cbFull,
R+   time = times,
R+   newdata = newx,
R+   s = "lambda.1se",
R+   method = "numerical"
R+ )
\end{CodeInput}
\end{CodeChunk}

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of coefficients with values greater than 0 are displayed in
\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
larger estimated coefficients. This is expected, as there are no
restrictions on the size of coefficients. The smallest coefficient
estimates are found in penalized case-base model, aside from cancer
state (cano).

@@

\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity. Shrinkage occurs due to penalization.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized case-base model compared to both Cox regression models. The
differences, however, are minimal. All three outperform the Kaplan-Meier
estimator.

@@

\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

In the previous case studies, we only considered covariates that were
fixed at baseline. In this next case study, we will use the Stanford
Heart Transplant data
\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
case-base sampling can also be used in the context of time-dependent
exposure. As an example that already appeared in the literature,
case-base sampling was used to study vaccination safety, where the
exposure period was defined as the week following vaccination
\citep{saarela2015case}. Hence, the main covariate of interest,
i.e.~exposure to the vaccine, was changing over time. In this context,
case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

As before, we can look at the population-time plot for a graphical
summary of the event incidence. Here, we coloured the exposed
person-time (i.e.~after transplant) in a darker shade of gray. As we can
see, most events occur early during the follow-up period, and therefore
we do not expect the hazard to be constant.

@@

@@

@@

\begin{CodeInput}
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+   data = cb_data, time = "futime"
R+ )
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+   data = cb_data, time = "futime"
R+ )
R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
R+   data = cb_data, time = "futime"
R+ )
\end{CodeInput}
\end{CodeChunk}

Note that the third model includes an interaction term between exposure
and follow-up time. In other words, this model no longer exhibit
proportional hazards. The evidence of non-proportionality of hazards in
the Stanford Heart Transplant data has been widely discussed
\citep{arjas1988graphical}.

@@

\begin{CodeInput}
R> plot(fit3, hazard.params = list(xvar = "futime",
R+                                 by = "exposure",
R+                                 alpha = 0.05,
R+                                 ylab = "Hazard",
R+                                 data = cb_data))
\end{CodeInput}

@@

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit3,
R+   newdata = newdata,
R+   time = seq(0, 425, length.out = 100)
R+ )
R> 
R> class(absrisk)
\end{CodeInput}

@@

\begin{CodeInput}
R> 
R> plot(absrisk,
R+      id.names = c("No Tx", "Tx")) + 
R+   ylab("Cumulative Incidence") +
R+   xlab("Follow-up time (in days)") +
R+   paper_gg_theme
\end{CodeInput}

@@

This report was generated on 2020-08-31 10:05:10 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
#> Head:     [0be28f4] 2020-08-25: switch to plot method for hazard
\end{CodeOutput}
\end{CodeChunk}

>>>>

\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}

@@

\toprule

@@

\bottomrule
\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{CoxRidge}        &                          & \checkmark                         & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\ 
\bottomrule

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                                event = "DeadOfPrCa", exposure = "ScrArm")
R> inherits(pt_object, "popTime")
\end{CodeInput}

@@

\begin{CodeInput}
R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
R> 
R> fits <- lapply(fmla, function(i) {
+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
+ })
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
R> new_time <- seq(0,14,0.1)
R> 
R> risks <- lapply(fits, function(i) {
+   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
+ })
\end{CodeInput}
\end{CodeChunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted above, the usual asymptotic results hold for
likelihood ratio tests built using case-base sampling models. Therefore,
we can easily test the null hypothesis that the exponential model is
just as good as the larger (in terms of number of parameters) spline
model:

@@

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
#> 1     54538       6052                         
#> 2     54535       5785  3      267   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
\centering
\begin{tabular}[t]{>{}lcc}
\toprule
Model & casebase::fitSmoothHazard & survival::survreg\\
\midrule
\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
\end{tabular}
\end{table}

@@

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
prostate cancer hazard ratio and 95\% confidence interval for screening
vs.~control group as a function of time using the \code{plot} method for
objects of class \code{singleEventCB}:

@@

\begin{CodeInput}
R> plot(fit_inter, type = "hr", newdata = new_data,
+      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+   data = bmtcrr
+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
+                         const(Phase) + const(Source) + const(Age),
+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. For more information
about this dataset, the reader is encouraged to look at the
documentation in our package.

In this section, we compare our penalized case-base approach to two
other models: Cox regression, and penalized Cox regression (using the
\pkg{glmnet} package). In all three models, we use all covariates except
\code{sps} and \code{aps}. These two variables correspond to scores for
predicting the outcome that were developed as part of the original
study.

To compare the performance of our models, we first divide the data into
a training and a test data, with 95\% of the observations used for
training all three models. To assess both discrimination and
calibration, we use a time-dependent version of the classical Brier
score that is adjusted for censoring \citep{graf1999ass}. The Brier
score can be used with both parametric and semi-parametric models. We
use the \pkg{riskRegression} to compute these scores for all models.

For our penalized case-base model, we opt for the natural log of time;
recall that this corresponds to a Weibull distribution. For fitting the
penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix
interface to the function \code{fitSmoothHazard}. In order to use this
interface, we first create a matrix \code{y} containing the time and
event variables, and a matrix \code{x} containing all remaining
covariates we would like to include in the model space. We used lasso
penalization by setting \code{alpha = 1}. Moreover, we use the argument
\code{penalty.factor} to give a regularization weight of 0 to the time
variable, and therefore force it into the final model.

@@

\begin{CodeInput}
R> # Create matrices for inputs
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> pen_cb <- casebase::fitSmoothHazard.fit(x, y,
+   family = "glmnet",
+   time = "d.time", event = "death",
+   formula_time = ~ log(d.time), alpha = 1,
+   ratio = 10, standardize = TRUE,
+   penalty.factor = c(0, rep(1, ncol(x)))
+ )
\end{CodeInput}
\end{CodeChunk}

Our first comparison plot examines the covariates that were selected
between all three hazard functions. In \ref{fig:cs3lolliPlot}, we show
the coefficient estimates for covariates that we selected by both
penalized Cox and penalized case-base. From this plot, we can clearly
see the shrinkage effect of using penalized estimation.

@@

\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. Only covariates common among all three models were retained. Shrinkage occurs due to penalization.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

Next, we compare the cumulative incidence estimation over the test set.
The probabilities over time for each observation are averaged, resulting
in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). We
can see some minimal differences between the three models, with the
Kaplan-Meier estimate giving a lower estimate than our three models. We
also note that the apparent smoothness of the Cox and penalized Cox
curves is due to the large number of observations in the training set,
which is used to derive the Breslow estimate of the baseline hazard. As
described above, we compare the performance between the models by
computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier}
(B), we can see that the Brier score is larger for the Kaplan-Meier
estimate than for the other three models. On the other hand, the
differences between these three models are minimal.

@@

\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

In brief, case-base sampling can be used in conjunction with penalized
logistic regression to perform variable selection with survival models.

@@

In the previous case studies, we only considered covariates that were
fixed at baseline. In this next case study, we use the Stanford Heart
Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
to show how case-base sampling can also be used in the context of
time-dependent exposure. This feature of case-base sampling has been
explored in the literature, in the context of vaccination safety
\citep{saarela2015case}. In this study, the exposure period was defined
as the week following vaccination. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

As above, we can look at the population-time plot for a graphical
summary of the event incidence. Here, we colour the exposed person-time
(i.e.~after transplant) in a darker shade of gray. As we can see, most
events occur early during the follow-up period, and therefore we do not
expect the hazard to be constant.

@@

\begin{CodeInput}
R> cb_data <- sampleCaseBase(jasa,
+   time = "futime",
+   event = "fustat", ratio = 100
+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
+   txtime = time_length(accept.dt %--% tx.date,
+                        unit = "days"
+   ),
+   exposure = case_when(
+     is.na(txtime) ~ 0L, # No transplant
+     txtime > futime ~ 0L,
+     txtime <= futime ~ 1L
+   )
+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+   data = cb_data, time = "futime"
+ )
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+   data = cb_data, time = "futime"
+ )
R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
+   data = cb_data, time = "futime"
+ )
\end{CodeInput}
\end{CodeChunk}

Note that the third model includes an interaction term between exposure
and follow-up time. In other words, this model no longer exhibits
proportional hazards. The evidence of non-proportionality of hazards in
the Stanford Heart Transplant data has been widely discussed
\citep{arjas1988graphical}.

@@

\begin{CodeInput}
R> plot(fit3, hazard.params = list(xvar = "futime",
+                                 by = "exposure",
+                                 alpha = 0.05,
+                                 ylab = "Hazard",
+                                 data = cb_data))
\end{CodeInput}

@@

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit3,
+   newdata = newdata,
+   time = seq(0, 425, length.out = 100)
+ )
R> 
R> class(absrisk)
\end{CodeInput}

@@

\begin{CodeInput}
R> 
R> plot(absrisk,
+      id.names = c("No Tx", "Tx")) + 
+   ylab("Cumulative Incidence") +
+   xlab("Follow-up time (in days)") +
+   paper_gg_theme
\end{CodeInput}

@@

This report was generated on 2020-09-15 17:46:19 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
#> Head:     [ddc5b48] 2020-09-14: Merge pull request #29 from sahirbhatnagar/editSahir
\end{CodeOutput}
\end{CodeChunk}
